---
title: "Analysis of Caribou Movements and Geospatial Covariates with the TuktuTools R Package"
author: "Eliezer Gurarie and Ophélie Couriot"
date: "`r Sys.Date()`"
output: html_document:
  number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Introduction

During this workshop, we will demonstrate the `TuktuTools` package, which has been developed for studying Tuktu (caribou) movement ecology and spatial patterns. 

To date, this package contains sets of functions that prep and filter movement data:

- `process_moveNWT` - process data from Movebank;
- `removeoutliers` - flags fixes considered as *outliers*;
- `prepData` - prepares and filters data to a specific period of time and/or a minimum number of fixes per day; 
- `scan_tracks` - visualizes individual paths through time

The package also contains functions that analyze spatial patterns, including:
- `getSpeed` - computes individual movement rates, displacements, speeds;
- `getDailyMean` - estimates individual mean daily location (i.e., the mean x and y coordinates of all the daily locations for a given individual);
- `getLoCoH` and `getKernelUD` - estimate ranging areas with Local Convex Hulls (LoCoH) or Kernel Utilization Distributions (KUD));
- `getPairwiseDistance` and `getPairwiseOverlap` - estimate daily pairwise distance or pairwise overlap between pairs of individuals;
- `estimateCalving` - estimate calving status (non-calving or calving, calving with survival or calving with calf death), calving timing, and calving location for given females, using movement rate; 
- `estimateMigration_stan` - estimates population-level migration timing using STAN

The package contains the following data:

- `caribou` - anonymized movement data for 4 individual Barren-ground caribou;
- `simulated_migrations` - simulated movement tracks for 18 individuals

# Installation

To install the current version of this package in R:

```{r, eval = FALSE}
install.packages("devtools")

devtools::install_github("ocouriot/TuktuTools", build_vignettes = TRUE)
```


# Chapter 1: Data and Vizualisation

## Data

The `TuktuTools` package contains movement data from 4 caribou individuals, which have been anonymized.
The package also contains data from 18 simulated tracks.

```{r}
require(TuktuTools)
data(package = 'TuktuTools')
data(caribou)
```

## Prepare and filter data



## Vizualise data

### Using `scan_track`

### Vizualisation on a map

```{r}
# transform the dataset in a simple feature object (projected object)
caribou_sf <- caribou %>% st_as_sf(coords=c("Lon", "Lat"), crs = 4326)

# download the map using the basemaps::basemap_raster
bbox <- st_bbox(caribou_sf) + c(-1,-1,1,1)
map <- basemaps::basemap_raster(ext = bbox, map_service = "osm", map_type = "topographic", map_res = 2)

# create lines for each individual during summer only after 2006
# transform in crs: 3857 (same as the map for plotting)
lines <- caribou_sf %>% 
  group_by(ID) %>% arrange(Time) %>%
  summarize(do_union = FALSE) %>%
  st_cast("LINESTRING") %>% st_transform(3857)
  
# add colors using viridis
library(viridis)
cols <- data.frame(ID = unique(caribou_sf$ID),
                   cols = viridis(length(unique(caribou_sf$ID))))
lines <- lines %>% merge(cols, by = "ID")

# Draw the map and add the individuals tracks with legend
raster::plotRGB(map, asp = 1, alpha = 100)
plot(st_geometry(lines), add = TRUE, col = alpha(lines$cols, alpha = .6), lwd = 2)
legend("right", legend = cols$ID, 
     col = cols$cols,
     lwd = 2,  ncol=1)
```



# Chapter 2: Basic analysis using TuktuTools

## Get the daily mean location

## Get the daily movement rate

## Get the individuals' Kernel Utilization Distribution

## Get the individuals' Local Convex Hull (LoCoH)


# Chapter 3: Estimating parturition time and location


# Chapter 4: Estimating migration with STAN
